Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-07-13 manuscript/manuscript.tex
Continue writing analytical solutions

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}
\noindent where
\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesseroid-tensor-kernel}
\end{equation}

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and
\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

@@

\iftwocol{
\begin{equation}
\begin{split}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx& \\
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    & W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\end{split}
\label{eq:glq-var-dens}
\end{equation}
}
{
\begin{equation}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

\iftwocol{
\begin{equation}
\begin{split}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'} -& \\
    \sqrt{r^2 + r'^2 - 2rr'}
    & \Big] \frac{r'\rho(r')}{r} \, dr'.
\end{split}
\label{eq:shell-pot-sqrts}
\end{equation}
}
{
\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'}  -
    \sqrt{r^2 + r'^2 - 2rr'} 
    \Big] \frac{r'\rho(r')}{r} \, dr'.
\label{eq:shell-pot-sqrts}
\end{equation}
}

For our purposes we can assume that the computation point $Q$ is outside the outer radii, i.e. $r>R_2$, and simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

@@

>>>>

\begin{equation}
    V(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \frac{\rho(r')}{\ell} \kappa \,  dr' d\phi' d\lambda',
\label{eq:tesseroid-pot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') \frac{\Delta_\alpha}{\ell^3}
    \kappa \, dr' d\phi' d\lambda',
\label{eq:tesseroid-grav}
\end{equation}
\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G
    \int\limits_{\lambda_1}^{\lambda_2}
    \int\limits_{\phi_1}^{\phi_2}
    \int\limits_{r_1}^{r_2}
    \rho(r') I_{\alpha\beta} \, \kappa \, dr' d\phi' d\lambda' ,
    \label{eq:tesseroid-tensor}
\end{equation}

\begin{equation}
    I_{\alpha\beta} =
    \left(
        \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
        \frac{\delta_{\alpha\beta}}{\ell^3}
    \right) ,
    \label{eq:tesseroid-tensor-kernel}
\end{equation}

\noindent $\alpha, \beta \in \{x, y, z\}$,
$\rho(r')$ is the density function that depends on the radius coordinate,
$\delta_{\alpha\beta}$ is Kronecker's delta,
$G = 6.674\times10^{-11}\, \text{m$^3$kg$^{-1}$s$^{-1}$}$ is the gravitational constant and

\begin{equation}
    \Delta_x = r'[\cos\phi\sin\phi' - \sin\phi\cos\phi'
               \cos(\lambda' - \lambda)],
\end{equation}
\begin{equation}
    \Delta_y = r' \cos \phi' \sin(\lambda' - \lambda),
\end{equation}
\begin{equation}
    \Delta_z = r' \cos \psi - r,
\end{equation}
\begin{equation}
    \kappa = {r'}^2 \cos \phi',
\end{equation}
\begin{equation}
    \ell = \sqrt{{r'}^2 + r^2 - 2 r' r \cos \psi},
\label{eq:ell}
\end{equation}
\begin{equation}
    \cos\psi = \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda).
\label{eq:cospsi}
\end{equation}

@@

\iftwocol{
\begin{equation}
    \begin{split}
        \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') 
        d\Omega \approx& \\
        A 
        \sum\limits_{i=1}^{N^r}
        \sum\limits_{j=1}^{N^\phi}
        \sum\limits_{k=1}^{N^\lambda}
        & W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
    \end{split}
\label{eq:glq-var-dens}
\end{equation}
}{
\begin{equation}
    \iiint\limits_\Omega \rho(r') g(r', \phi', \lambda') d\Omega \approx
    A 
    \sum\limits_{i=1}^{N^r}
    \sum\limits_{j=1}^{N^\phi}
    \sum\limits_{k=1}^{N^\lambda}
    W_i^r W_j^\phi W_k^\lambda \rho(r_i) g(r_i, \phi_j, \lambda_k),
\label{eq:glq-var-dens}
\end{equation}
}

@@

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}(r) = 2\pi G
        \int\limits_{R_1}^{R_2}
        \Big[ & \sqrt{r^2 + r'^2 + 2rr'} - \\
        & \sqrt{r^2 + r'^2 - 2rr'}
        \Big] \frac{r'\rho(r')}{r} \, dr'.
    \end{split}
\label{eq:shell-pot-sqrts}
\end{equation}
}{
\begin{equation}
    V_\text{sh}(r) = 2\pi G
    \int\limits_{R_1}^{R_2}
    \Big[ \sqrt{r^2 + r'^2 + 2rr'}  -
    \sqrt{r^2 + r'^2 - 2rr'} 
    \Big] \frac{r'\rho(r')}{r} \, dr'.
\label{eq:shell-pot-sqrts}
\end{equation}
}

For our purposes we can assume that the computation point $Q$ is outside the outer radii, i.e. $r>R_2$ ($R_2 \geq r'$), and simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

@@

For any density $\rho(r')$, the gradient and Marussi tensor components can be obtained as \citet{Grombein2013} does. Only the vertical gradient component ($g_z$) and the diagonal Marussi tensor components ($g_{xx}$, $g_{yy}$, $g_{zz}$) are not zero, and can be easily determined from the potential $V_\text{sh}$:

\begin{equation}
    g_z = \frac{V_\text{sh}(r)}{r},
\end{equation}
\begin{equation}
    g_{xx} = g_{yy} = -\frac{V_\text{sh}(r)}{r^2}, \quad
    g_{zz} = \frac{2V_\text{sh}(r)}{r^2}.
\end{equation}

For practical purposes we are going to obtain the gravity potential for two very different functions in the perspective of they variation in $r'$: a linear density and an exponential density.

\subsubsection{Linear Density}

If the density of the spherical shell is a linear function in the radius coordinate, i.e.:

\begin{equation}
    \rho(r') = ar' + b,
\end{equation}

\noindent then the gravity potential generated on any computation point at distance $r$ from its centre can be obtained from the following expression:

\begin{equation}
    V_\text{sh}^\text{lin}(r) = \pi G \left[ 
    a \frac{R_2^4 - R_1^4}{r} +
    b \,\frac{4}{3} \frac{R_2^3 - R_1^3}{r} \right].
    \label{eq:shell-pot-linear}
\end{equation}

The second term on equation \ref{eq:shell-pot-linear} reassembles the potential generated by a spherical shell with homogeneous density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
So, this potential can be read as the combination of a spherical shell with the later homogeneous density and the same shell with variable density $\rho(r') = ar'$.

\subsubsection{Exponential Density}

In case of a spherical shell with exponential density in the radius coordinate,

\begin{equation}
    \rho(r') = Ae^\frac{r'-\Delta h}{b},
\end{equation}

\noindent we have used \citet{sagemath} to obtain the following expression for the potential:

\iftwocol{
\begin{equation}
    \begin{split}
        V_\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
        Ab \, e^{-\frac{\Delta h}{b}} \Big[ 
        & (R_2^2 - 2R_2 b + 2b^2)e^\frac{R_2}{b} - \\
        & (R_1^2 - 2R_1 b + 2b^2)e^\frac{R_1}{b} \Big].
    \end{split}
\end{equation}
}{
\begin{equation}
    V_\text{sh}^\text{sh}^\text{exp}(r) = \frac{4\pi G}{r} 
    Ab \, e^{-\frac{\Delta h}{b}} \left[ 
    (R_2^2 - 2R_2 b + 2b^2)e^\frac{R_2}{b} - 
    (R_1^2 - 2R_1 b + 2b^2)e^\frac{R_1}{b} \right].
\end{equation}
}
